---
title: "Clustering"
author: "M. van Vugt"
date: "01/01/2020"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
---

# Setup
We will clean the environment, setup the locations and create a datestamp.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```


_Clean the environment and set datestamp_
```{r echo = FALSE}
rm(list = ls())

Today <- format(as.Date(as.POSIXlt(Sys.time())), "%Y%m%d")
Today_Report <- format(as.Date(as.POSIXlt(Sys.time())), "%A, %B %d, %Y")

```


_Set locations and working directories_
```{r LocalSystem, echo = FALSE}

### Change these paths accordingly, without the backslash at the end!
ROOT_loc <- "/Volumes/home"
PROJECT_loc <- paste0(ROOT_loc, "/Side-projects/ReproClus/repro-clus")

# If using the cookiecutter set-up, these should be fine, otherwise change accordingly
DATA_in <- paste0(PROJECT_loc, "/data/raw")
DATA_temp <- paste0(PROJECT_loc, "/data/temp")
DATA_out <- paste0(PROJECT_loc, "/data/processed")

CONFIG_loc <- paste0(PROJECT_loc, "/config")

OUT_loc <- paste0(PROJECT_loc, "/results/output")
FIG_loc <- paste0(PROJECT_loc, "/results/figures")

```


_A package-installation function_
```{r Function: installations}

library(docstring)

install.packages.auto <- function(x) {
  #' @title Install and load packages
  #' @description This function installs and loads packages automatically.
  #' @param x The name of the package, using quotes.
  #' @return A text message specifying whether installing and loading of the package 
  #' was successful.
  
  x <- as.character(substitute(x))
  if (isTRUE(x %in% .packages(all.available = TRUE))) { 
    eval(parse(text = sprintf("require(\"%s\")", x)))
  } else { 
    # Update installed packages - this may mean a full upgrade of R, which in turn
    # may not be warrented. 
    # update.packages(ask = FALSE) 
    eval(parse(text = sprintf("install.packages(\"%s\", dependencies = TRUE, repos = \"http://cran-mirror.cs.uu.nl/\")", x)))
  }
  if (isTRUE(x %in% .packages(all.available = TRUE))) { 
    eval(parse(text = sprintf("require(\"%s\")", x)))
  } else  {
    source("http://bioconductor.org/biocLite.R")
    # Update installed packages - this may mean a full upgrade of R, which in turn
    # may not be warrented.
    #biocLite(character(), ask = FALSE) 
    eval(parse(text = sprintf("biocLite(\"%s\")", x)))
    eval(parse(text = sprintf("require(\"%s\")", x)))
  }
  return(paste0(x, " is installed and loaded correctly"))
}
```



_Load those packages_
```{r loading_packages}

install.packages.auto("readr")
install.packages.auto("dplyr")
install.packages.auto("tableone")
install.packages.auto("haven")
install.packages.auto("poLCA")
install.packages.auto("ggplot2")
install.packages.auto("data.table")
install.packages.auto("reshape2")

```


## Cohort characteristics

Before starting analysis, we want to have an overview of the cohort characteristics. To that end, a table is created including the overall and per-cohort characteristics of the variables available so far. 

_Load data and create categorical variables for Age and BMI_
```{r Load data, include=FALSE}

# Start with the data created for this script (Dummy data in this example)
data <- read_delim(paste0(DATA_in, "/data.csv"), ";",
                   escape_double = FALSE, 
                   col_types = cols(Sex = col_factor(levels = c("M","F")), 
                                    Hypertension = col_factor(levels = c("yes", "no")), 
                                    Isch = col_factor(levels = c("yes", "no")), 
                                    HFrEF = col_factor(levels = c("HFrEF", "HFpEF")), 
                                    CABG = col_factor(levels = c("yes", "no")), 
                                    Diabetes = col_factor(levels = c("yes", "no")),
                                    Cohort = col_factor(levels = c("Cohort1", "Cohort2"))), 
                   trim_ws = TRUE)
str(data)

# The latent class analysis only accepts categorical data, 
# so convert the age and BMI into categories
data$Age_cat <- as.factor(ifelse(data$Age <= 60, "<60",
                                 ifelse(data$Age >= 76, ">75", "60-75")))
data$BMI_cat <- as.factor(ifelse(data$BMI <= 25, "<25",
                                 ifelse(data$BMI >= 30, "≥30", "25-29")))

str(data)
```

_Create variables for table_
```{r Create variables for table, include = FALSE}

names(data)

# For the Characteristics table, we are going to make vectors for the data to be in the table, 
# the data to be treated as categorical and nonnormal.
myVars <- c("Cohort", "Age", "Age_cat", "Sex", "BMI", "BMI_cat", "HFrEF", "Isch",
            "Hypertension", "Diabetes", "CABG")
catVars <- c("Cohort", "Sex", "Age_cat", "BMI_cat", "HFrEF", "Isch", "Hypertension", 
             "Diabetes", "CABG")
nnorVars <- c("BMI")

```

_Make characteristics table_
```{r Cohort tables, echo=FALSE, warning=FALSE}

tblcohort <- CreateTableOne(vars = myVars, data = data, strata = "Cohort", 
                            addOverall = TRUE, factorVars = catVars)

ex <- print(tblcohort, showAllLevels = TRUE, missing = TRUE, nonnormal = nnorVars,
      formatOptions = list(big.mark = ","), quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
ex
write.csv(ex, paste0(OUT_loc, "/", Today, "Cohorts_table.csv")) 

rm(tblcohort, ex)
```

_Create final dataframe_
```{r Setting up dataframe, echo=FALSE}

# This dataframe is created to contain only the variables that are going to be used in the 
# Latent Class Analysis. For the dummy data, that are all variables, but in case you would
# load data with more variables, this can be changed.
LCA_cohort <- data[, c("ID","Sex", "Age", "Age_cat", "BMI_cat", "HFrEF", "Isch",
                      "Hypertension", "CABG", "Diabetes")]
    
LCA_cohort
    
summary(LCA_cohort)

```

##  Performing Latent Class analysis

First, several models are tried, using 2-10 classes and repeating each model 10 times to search for global log-likelihoods. The results of each model are stored, so we can visually evaluate the performance, but the best model is already determined by looking at the BIC.
For now, this model contains: Sex, Age_cat, HFrEF, Isch, Hypertension, CABG, Diabetes.

```{r LCA, include=FALSE}

# Specify model, specifying which variables will be included.
f <- with(LCA_cohort, 
          cbind(Sex, Age_cat, BMI_cat, HFrEF, Isch, Hypertension, CABG, Diabetes)~1)
min_bic <-  1000000

result <- vector("list", 3)  # Create vector to store results of for-loop in      
for (nclass in 2:10) {
  lc <- poLCA(f, LCA_cohort, nclass = nclass, maxiter = 50000,
              tol = 1e-5, na.rm = F, nrep = 10, verbose = T, calc.se = T) 
  result[[nclass]] <- lc
  if(lc$bic < min_bic){
    min_bic <- lc$bic
    LCA_best_model <- lc
      }
}

lca2 <- result[[2]]
lca3 <- result[[3]]
lca4 <- result[[4]]
lca5 <- result[[5]]
lca6 <- result[[6]]
lca7 <- result[[7]]
lca8 <- result[[8]]
lca9 <- result[[9]]
lca10 <- result[[10]]

LCA_best_model

LCA_best_model$predclass # predicted in which class patient falls
LCA_best_model$posterior # probability of each patient for the classes
rm(f, result)
```


## Visualization choosing the best model

TABLE and PLOT showing how to select best model.

```{r Table and plot, echo=FALSE}
    
Classes <- c(2, 3, 4, 5, 6, 7, 8, 9, 10)
AIC <- c(lca2$aic, lca3$aic, lca4$aic, lca5$aic, 
         lca6$aic, lca7$aic, lca8$aic, lca9$aic, lca10$aic)

BIC <- c(lca2$bic, lca3$bic, lca4$bic, lca5$bic, 
         lca6$bic, lca7$bic, lca8$bic, lca9$bic, lca10$bic)

Likelihoodratio <- c(lca2$llik, lca3$llik, lca4$llik, lca5$llik, 
                     lca6$llik, lca7$llik, lca8$llik, lca9$llik, lca10$llik)

Chi2 <- c(lca2$Chisq, lca3$Chisq, lca4$Chisq, lca5$Chisq, 
          lca6$Chisq, lca7$Chisq, lca8$Chisq, lca9$Chisq, lca10$Chisq)
    
df <- data.frame(Classes, AIC, BIC, Likelihoodratio, Chi2)
df <- melt(df, id.vars = c("Classes"))
df$class <- as.factor(df$Classes)
df
    
fit_plot <- ggplot(df) + geom_point(aes(x = Classes, y = value)) +
                        geom_line(aes(Classes, value, group = 1)) +
                        facet_wrap(~variable, scales = "free") +
                        theme_bw()
pdf(paste0(FIG_loc, "/", Today, "_Model_selection.pdf"))
fit_plot
dev.off()

fit_plot
rm(Classes, AIC, BIC, Likelihoodratio, Chi2, lca2, lca3, lca4, lca5, lca6, lca7, lca8, lca9, lca10, df, nclass, min_bic, lc)

```
    
## Looking at the best model

For now, the model with 3 classes is best and therefore we'll continue with this one. (For the example data, this is not the case, I am sorry!).
Below the model is shown, together with the percentages and numbers of patients belonging to each class.

```{r Model with 3 classes, echo=FALSE}

lca3.polca <- poLCA(cbind(Sex, Age_cat, BMI_cat, HFrEF, Isch, Hypertension, CABG, Diabetes)~1, 
                       data = LCA_cohort,
                       nclass=3,
                       na.rm=F,
                       graphs=F,
                       maxiter = 50000,
                       nrep = 10)

predclass <- data.table(lca3.polca$predclass)
setnames(predclass, old = "V1", new = "predclass")
    
lca <- cbind(data, predclass)
    
lapply(prop.table(table(lca$predclass))*100, round, 1) 
    
table(lca$predclass)

```

The chunk above adds a column which states to what class each patient belongs. It also shows how many patients (both in number and percentages) belong to each class.

```{r Plots based on age and sex, include=FALSE}

probability <- data.table(lca3.polca$posterior)
setnames(probability, old = c("V1", "V2", "V3"), new = c("p1", "p2", "p3"))
    
lca <- cbind(lca, probability)
    
rm(probability, predclass)

lca3.plots <- lca[, c("Age_cat", "Sex" , "p1", "p2", "p3")]
lca3.plots

probs <- lca3.plots %>%
  group_by(Age_cat, Sex) %>%
  summarize_all(list(mean = mean))
probs

probs <- melt(probs, vars = c("Age_cat", "Sex"))
probs <- na.omit(probs)
probs <- data.table(probs)
probs

setnames(probs, old = c("Age_cat", "variable"), new = c("Age", "Class"), skip_absent = T)
levels(probs$Class)
levels(probs$Class) <- c("Probability class 1", "Probability class 2", 
                             "Probability class 3")

probs$Age <- factor(probs$Age, levels = c("<60", "60-75", ">75"))   

sex_lca <- ggplot(probs, aes(x = Age, y = value, 
                             group = interaction(Class, Sex), 
                             color = Class, linetype = Sex)) +
  geom_line(size = 1.2) +
  scale_colour_manual(values = 
                        c("#D55E00", "#56B4E9",  "#0072B2","#009E73", "#E69F00", "#999999")) +
  labs(x = "Age (in three categories)",y="Probability in class") +
  theme_bw() + theme(text = element_text(size=20), 
                     axis.text.x = element_text(angle=45, hjust=1))

pdf(paste0(FIG_loc, "/", Today, "_sex_age_classes.pdf"))
sex_lca 
dev.off()

rm(probs)
```

```{r Show plot, echo=FALSE}

sex_lca 

```


## Create new table

```{r Charasteristics table of classes, include=FALSE}

tblclass <- CreateTableOne(vars = myVars, data = lca, strata = "predclass", 
                            addOverall = TRUE, factorVars = catVars)

# Print table and export it into csv to easily copy it to Word
ex_class <- print(tblclass, showAllLevels = TRUE, missing = TRUE, nonnormal = nnorVars,
                  formatOptions = list(big.mark = ","), quote = FALSE, noSpaces = TRUE,
                  printToggle = FALSE)

write.csv(ex_class, paste0(OUT_loc, "/", Today, "_Class_table.csv")) 
rm(tblclass, ex_class)
```

## Output processed data

```{r Output data, include = FALSE}

write_delim(lca, paste0(DATA_out, "/", Today, "_LCA_data.csv"))

```
